<<<<<<< HEAD
---
title: "Questions: Rearranging equations involving trigonometry and logarithms"
author: "Ellie Gurini"
abstract-title: Summary
abstract: This is a question set based on the content on the Guide, Introduction to rearranging equations involving trigonometry and logarithms. It's intended to be completed after having read that, and there is an answer sheet to match this question set. Note that answers are vague, as questions are based off of examples in the guide.

---

```{r, setup, include = FALSE}
library("webexercises")
```

*Before attempting these questions, it is highly recommended that you read [Guide: Introduction to rearranging equations with trig and logs] (rearrangingtrigandlogs.qmd)*

## Q1 {-}

Solve each of these equations assuming we are using radians as our angle unit.

1.1 $sin(x)=\frac{\sqrt2}{2}$ 
1.2 $cos(2x+1)=\frac{1}{2}$
1.3 $tan(5x-1)=\frac{\sqrt2}{2}$
1.4 $cos(x^2+4x+3)=1$


=======
---
title: "Questions: Rearranging equations involving trigonometry and logarithms"
author: "Ellie Gurini"
abstract-title: Summary
abstract: This is a question set based on the content in the guide Introduction to rearranging equations involving trigonometry and logarithms. It's intended to be completed after having read that, and there is an answer sheet to match this question set. Note that answers are vague, as questions are based off of examples in the guide.

---

```{r, setup, include = FALSE}
library("webexercises")
```

*Before attempting these questions, it is highly recommended that you read the guide 'Introduction to rearranging equations with trig and logs' *

## Q1 {-}

Solve each of these equations assuming we are using radians as our angle unit.

1.1 $sin(x)=\frac{\sqrt2}{2}$ 
1.2 $cos(2x+1)=\frac{1}{2}$
1.3 $tan(5x-1)=\frac{\sqrt2}{2}$
1.4 $cos(x^2+4x+3)=1$

## Q2 {-}

Using that $1+\cot^2(x)=\csc^2(x)$, prove that $\sin^2(x)+\cos^2(x)=1$.


>>>>>>> refs/remotes/origin/14--rearranging-equations-involving-trig-and-logs
