---
title: "Answers: Introduction to logic"
author: "Jessica Taberner"
abstract-title: Summary
abstract: Answers to questions relating to the guide on introduction to logic.
draft: true
---

```{r, setup, include = FALSE}
library("webexercises")
```

*These are the answers to [Questions: Introduction to logic.](../questions/qs-introtologic.qmd)*

**Please attempt the questions before reading these answers!**

## Q1 Propositions{-}

**1.1.** "Boole's Burgers is a burger restaurant" is a proposition.

**1.2.** "Our burgers taste amazing!" is not a proposition.

**1.3.** “All burgers in the restaurant cost more than $£2$” is a proposition.

**1.4.** “Please give me a cheeseburger” is not a proposition.

**1.5.** “What are your burger toppings?” is not a proposition.

**1.6.** “Some customers prefer vegan burgers” is a proposition.

**1.7.** “Are the fries sold out?” is not a proposition.

**1.8.** “The fries are sold out” is a proposition.

## Q2 Connectives{-}

**2.1.** The logical structure of "If you buy two burgers, you'll get a discount" is $P \rightarrow Q$.

**2.2.** The logical structure of "I'll get one cheeseburger and one vegan burger" is $P \wedge Q$.

**2.3.** The logical structure of "I won't get cheese on my burger" is $\neg P$.

**2.4.** The logical structure of "I'll get a hamburger, or I'll get a milkshake" is $P \vee Q$.

**2.5.** The logical structure of "I'll only get a cheeseburger if you get one too" is $P \leftrightarrow Q$.

**2.6.** The logical structure of "If you go to Boole's Burgers, I'll go too, and I'll get a vegan burger" is $P \rightarrow (Q \wedge R)$.

## Q3 Truth tables{-}

**3.1.** $P \rightarrow Q$

```{r, echo=FALSE}
library(knitr)

df <- data.frame(
  `$P$` = c("T", "T", "F", "F"),
  `$Q$` = c("T", "F", "T", "F"),
  `$P \\rightarrow Q$`= c("T", "F", "T", "T"),
  check.names = FALSE
)

kable(df, format = "markdown", escape = FALSE)
```

**3.2.** $\neg P$

```{r, echo=FALSE}
library(knitr)

df <- data.frame(
  `$P$` = c("T", "F"),
  `$\\neg P$`= c("F", "T"),
  check.names = FALSE
)

kable(df, format = "markdown", escape = FALSE)
```

**3.3.** $(P \vee Q) \wedge \neg Q$

```{r, echo=FALSE}
library(knitr)

df <- data.frame(
  `$P$` = c("T", "T", "F", "F"),
  `$Q$` = c("T", "F", "T", "F"),
  `$(P \\vee Q) \\wedge \\neg Q$`= c("F", "T", "F", "F"),
  check.names = FALSE
)

kable(df, format = "markdown", escape = FALSE)
```

**3.4.** $\neg (P \leftrightarrow Q)$

```{r, echo=FALSE}
library(knitr)

df <- data.frame(
  `$P$` = c("T", "T", "F", "F"),
  `$Q$` = c("T", "F", "T", "F"),
  `$\\neg (P \\leftrightarrow Q)$`= c("F", "T", "T", "F"),
  check.names = FALSE
)

kable(df, format = "markdown", escape = FALSE)
```
 
**3.5.** $\neg (P\wedge Q)$

```{r, echo=FALSE}
library(knitr)

df <- data.frame(
  `$P$` = c("T", "T", "F", "F"),
  `$Q$` = c("T", "F", "T", "F"),
  `$\\neg (P \\wedge Q)$`= c("F", "T", "T", "T"),
  check.names = FALSE
)

kable(df, format = "markdown", escape = FALSE)
```

**3.6.** $(P \rightarrow Q) \rightarrow R$

```{r, echo=FALSE}
library(knitr)

df <- data.frame(
  `$P$` = c("T", "T","T", "T", "F", "F","F", "F"),
  `$Q$` = c("T", "F", "T", "F","T", "F", "T", "F"),
  `$R$` = c("T", "T", "F", "F", "T", "T", "F", "F"),
  `$(P \\rightarrow Q)\\rightarrow R$`= c("T", "T", "F", "T", "T","T", "F", "F"),
  check.names = FALSE
)

kable(df, format = "markdown", escape = FALSE)
```

**3.7.** $\neg (P \vee (Q \vee R))$

```{r, echo=FALSE}
library(knitr)

df <- data.frame(
  `$P$` = c("T", "T","T", "T", "F", "F","F", "F"),
  `$Q$` = c("T", "F", "T", "F","T", "F", "T", "F"),
  `$R$` = c("T", "T", "F", "F", "T", "T", "F", "F"),
  `$\\neg (P \\vee (Q \\vee R))$`= c("F", "F", "F", "F","F", "F", "F", "T"),
  check.names = FALSE
)

kable(df, format = "markdown", escape = FALSE)
```

**3.8.** $(P \leftrightarrow Q) \vee (\neg R)$

```{r, echo=FALSE}
library(knitr)

df <- data.frame(
  `$P$` = c("T", "T","T", "T", "F", "F","F", "F"),
  `$Q$` = c("T", "F", "T", "F","T", "F", "T", "F"),
  `$R$` = c("T", "T", "F", "F", "T", "T", "F", "F"),
  `$(P \\leftrightarrow Q) \\vee (\\neg R)$`= c("T", "F", "T", "T","F", "T", "T", "T"),
  check.names = FALSE
)

kable(df, format = "markdown", escape = FALSE)
```

## Q4 Equivalence{-}


**4.1.** $\neg (P \wedge Q)$ and $(\neg P) \vee (\neg Q)$ are logically equivalent. This identity is known as **De Morgan's Law**.

**4.2.** $P \rightarrow Q$ and $Q \rightarrow P$ are not logically equivalent.

**4.3.** $P \rightarrow Q$ and $(\neg P) \vee Q$ are logically equivalent.

**4.4.** $P \rightarrow Q$ and $\neg (P \vee Q)$ are not logically equivalent.

## Version history and licensing {-}

v1.0: initial version created 03/26 by Jessica Taberner as part of a University of St Andrews VIP project.

[This work is licensed under CC BY-NC-SA 4.0.](https://creativecommons.org/licenses/by-nc-sa/4.0/?ref=chooser-v1)
